### 2018-01-26 submission/manuscript.Rmd
discussion and conclusion -- ada suggestions

***Select Genera Models can Recapitulate Whole Community Models:***
Since specific genera increased RR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera based on RR and P-value significance and tested this using two approaches. The first approach used genus level data and tested whether there were any differences in AUC between all genera and selected genera during training on one study or testing on all other studies. The second approach used OTU level data and tested whether there was a generalized decrease in the 10-fold cross validation (CV) model AUC across studies when comparing all OTUs versus only OTUs that taxonomically classified to selected genera.

For the genera-based models the training set median AUC for model classification was similar for select genera and full genera models, for both tissue and stool studies [Figure S2-S3]. With respect to the test sets, comprised of genera data from other studies, both all and select genera models had a similar ability to detect adenomas or carcinomas, with the select genera models performing better in some instances [Figure S4-S6]. Conversely, the OTU-based models showed a slight decrease in median AUC between the full and select models, with one exception being carcinoma models for matched tissue [Figure 4 & 5].

In stool, the most common genera in the full genera-based models belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B]. Although all mouth-associated genera were present in carcinoma, none were present in the majority of studies and *Fusobacterium* was the only genus present in adenoma [Figure 6A & B]. For the full OTU-based models, *Ruminococcaceae* was present in the top 10 consistently for both adenoma and carcinoma models while *Roseburia* was only present in many adenoma models and *Bacteroides* was present in the overwhelming majority of the carcinoma models [Figure 6C & 6D].

Unlike stool, the tissue full genera-based Random Forest models showed no consistent representation of *Ruminococcaceae*, *Ruminococcus*, *Bacteroides*, and *Roseburia* in the top 10 across study [Figure S7]. The vast majority of the top 10 genera and OTUs occured in a study specific manner for these tissue based Random Forest models. For both the full genera and OTU Random Forest models of adenoma and carcinoma, there appears to be very little overlap in the top 10 most important variables between stool and tissue [Figure 6 & S7]. This discordance between stool and tissue also applies to the mouth-associated genera with one noticeable skew being that *Fusobacterium* and *Fusobacteriaceae* occur more often in the top 10 of Random Forest models using matched versus unmatched tissue samples [Figure S7B-C & S7E-F].

***CRC Studies are Underpowered for Detecting Small Effect Sizes:***
Based on our reported results we assessed whether the studies analyzed were realistically powered to identify small, medium, and large scale differences between cases and controls. When assessing the power of each study at different effect sizes the majority of studies, for both adenoma and carcinoma, acheived 80% power to detect a 30% or greater difference between groups [Figure 7A & B]. No study that was analyzed had the standard 80% power to detect an effect size difference that was equal to or below 10% [Figure 7A & B].  In order to achieve a power of 80%, for small effect sizes, studies used in our meta-analysis would need to recruit over 1000 individuals for each arm [Figure 7C].

@@

Our study identifies clear differences in diversity, both at the community level and for individual genera, that are present in patients with and without CRC [Figure 1-3]. Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of lesion. Even though mouth-associated genera increased the RR of carcinoma, they did not consistently increase the RR of adenoma. These mouth-associated genera are clearly important to carcinoma classification but our observations suggest that accounting for other community members can increase the ability of models to make predictions.

The data presented herein supports the importance of select genera for carcinoma but not necessarily adenoma formation. Our observations show that when only using the select genera to create models for carcinoma the AUC are similar to using all genera or OTUs [Figure 4 & 5]. This suggests that an interplay between a select number of potentially protective and exacerbating microbes, within the GI community, are crucial for carcinoma formation. Importantly, it suggests that there may be key members of the GI community that might be studied further to potentially reduce the risk of carcinoma. Conversely, using the present data, it is clear that new approaches may need to be used to identify members of the community that are associated with adenoma. Regardless of sample type and whether a full or select model was used our Random Forest models consistently performed poorly. Yet, the step-wise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1]. This change in diversity, at this early stage of disease, could be focal to the adenoma itself. One potential hypothesis from these observations is that at early stages of the disease, how the host interacts with these subtle changes is what ultimately leads to a thoroughly dysfunctional community that is supportive of CRC genesis.

Within stool, common GI microbes were most consistently present in the top 10 genera or OTUs across study [Figure 6]. Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be disciminative across the different studies for both adenoma and carcinoma [Figure 6]. This data would suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* is not as important as how these bacteria interact with the changing resident community. Based on these observations, it is possible to hypothesize that small changes in community structure lead to new niches in which any one of the mouth-associated genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma.

The tissue studies did not provide a clearer understanding of how the microbiota may be associated with lesion. Generally, the full OTU-based models of unmatched and matched [Figure S7E & F] tissue samples were concordant with stool showing that resident microbes were the most prevalent in the top 10 across study. Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 of the full CRC models [Figure S7B-C & E-F]. The majority of the tissue results seem to be study specific with many top 10 taxa being present only in a single study. Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models is worrying. The low bacterial biomass of tissue samples coupled with potential contamination could ultimately explain why the these results seem to be more sporadic than the stool results.

One important caveat to this study is that even though specific genera associated with species such as *Bacteroides fragilis* and *Streptococcus gallolyticus* subsp. *gallolyticus* were not identified, it does not necessarily mean these specific species are not important in human CRC [@bfrag_Boleij_2014; @strep_Aymeric_2017]. Since we are limited in our aggregation of the data, across study, to the genus level, it is not possible to clearly delineate what specific species are contributing to overall disease progression. Our observations are not inconsistent with the previous literature on either *Bacteroides fragilis* or *Streptococcus gallolyticus* subsp. *gallolyticus*. As an example, for stool, the full RF models consistently identified the genus *Bacteroides* as well as OTUs that classified as *Bacteroides* to be important model components across studies. Suggesting that even though *Bacteroides* may not increase the RR of lesion and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC. Additionaly, *Streptococcus gallolyticus* subsp. *gallolyticus* is a mouth-associated micobe and our study clearly shows that regardless of whether the sample is tissue or stool, mouth-associated genera are commonly associated with an increased RR of both adenoma and carcinoma.

The associations between the microbiota and adenoma are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes. None of the studies analyzed were properly powered to detect a 10% or lower change between cases and controls. A small effect size may well be the scope in which differences consistently occur in adenoma based on the results observed within our meta-analysis. Future studies investigating adenoma and the microbiota need to take power into consideration if we are to reproducibly study whether the microbiota contributes to polyp formation. In contrast to adenoma, our observations suggest that most studies analyzed have sufficient power to detect many changes in carcinoma because of large effect size differences between cases and controls [Figure 7].

@@

By aggregating together a large collection of studies from both feces and tissue, we are able to provide evidence in support of the importance of the bacterial community in CRC. Further, the data presented here suggests that mouth-associated microbes can gain a foothold within the colon and are are commonly associated with the greatest RR of carcinoma. Conversely, no conclusive signal with these mouth-associated micobes could be detected for adenoma. Our observations also highlight the importance of power and sample number considerations when undertaking investigations into the microbiota and adenoma due to the subtle changes in the community. Overall, the microbiota associations with carcinoma are much stronger than the associations with adenoma.

>>>>

***Select Genera Models can Recapitulate Whole Community Models:***
Since specific genera increased RR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera. We selected these genera based on their RR and P-value significance and used two approaches to test this question.
The first approach used genus level data and tested for differences in AUC between all genera and selected genera. A single study was used for training the model prior to testing on all other studies and this was repeated for every study in the meta-analysis. The second approach used OTU level data and tested for a generalized decrease in the 10-fold cross validation (CV) model AUC which is a common approach used to guard against over-fitting. This was applied across study and the AUC of the all OTUs model was compared against the model that used only OTUs that taxonomically classified to selected genera.

For the first approach using the genera-based models, the training set median AUC for model classification was similar for both the select genera and full genera models, for both tissue and stool studies [Figure S2-S3]. When analyzing the tests sets that were comprised of genera data from other studies, both models had a similar ability to detect individuals with adenomas or carcinomas, with the select genera models performing better in some instances [Figure S4-S6]. Conversely, the second approach that used OTU-based models showed a slight decrease in median AUC between the full and select models, with one exception to this generalization being the carcinoma models for matched colon tissue [Figure 4 & 5].

In stool, the most common genera in the top 10 most imporant variables, in models for the first approach, that used full genera-based models were *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B]. Regardless of sample type, mouth-associated genera were present in models for the carcinoma stage of CRC [Figure 6A & B]. Yet, none were present in the majority of studies and *Fusobacterium* was the only genus present in the adenoma stage of CRC [Figure 6A & B]. For the second approach that utilized full OTU-based models, *Ruminococcaceae* was present in the top 10 consistently for both adenoma and carcinoma models while *Roseburia* was only present in many adenoma models and *Bacteroides* was present in the overwhelming majority of the carcinoma models [Figure 6C & 6D].

Unlike the stool-based Random Forest models, the tissue-based models, for the full genera from the first approach, showed no consistent representation of *Ruminococcaceae*, *Ruminococcus*, *Bacteroides*, and *Roseburia* in the top 10 most important model variables across study [Figure S7]. The vast majority of the top 10 model variables for the genera- and OTU-based models using colon tissue tended to be study specific. Further, there was very little overlap in the top 10 important variables between adenoma and carcinoma stage models, regardless of whether colon tissue or stool was used  [Figure 6 & S7]. This discordance between stool and colon tissue samples also applies to the mouth-associated genera with one noticeable skew being that *Fusobacterium* and *Fusobacteriaceae* occur more often in the top 10 of matched versus unmatched colon tissue Random Forest models [Figure S7B-C & S7E-F]. This suggests that either the colon tissue microbiota is study and person dependent or that kit and/or other types of contamination associated with low biomass samples may be skewing the results.

***CRC Studies are Underpowered for Detecting Small Effect Sizes:***
Next, we assessed how much confidence should be placed in the reported outcomes from each individual study by calculating the ability to detect a difference (power) and sample size needed for small, medium, and large effect size differences between cases and controls. When assessing the power of each study at different effect sizes the majority of studies acheived 80% power to detect a 30% or greater difference between groups [Figure 7A & B]. No study that we analyzed had the standard 80% power to detect an effect size difference equal to or below 10% [Figure 7A & B].  In order to achieve a power of 80%, for small effect sizes, studies used in our meta-analysis would need to recruit over 1000 individuals for both the case and control arms [Figure 7C]

@@

Our study identifies clear differences in diversity, both at the community level and for individual genera, present in patients with and without CRC [Figure 1-3]. Although there was a step-wise decrease in diversity as disease progressed from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of CRC. Even though mouth-associated genera increased the RR of having a carcinoma, they did not consistently increase the RR of having an adenoma. Additionally, our observations suggest that by combining mouth-associated and CRC protective microbes we can classify either adenoma or carcinoma stage of disease as well as models that use the full community.

The data presented herein support the importance of select genera for carcinoma, but not necessarily adenoma, formation.
The results that we have presented show that the both genera and OTU select and full models, for the carcinoma stage of CRC, had similar AUCs [Figure 4 & 5]. This suggests that an interplay between a select number of potentially protective and exacerbating microbes within the GI community is crucial for carcinoma formation. Importantly, it suggests that there may be key members of the GI community that might be studied further to potentially reduce the risk of carcinoma. Conversely, using the present data, it is clear that new approaches may be needed to identify members of the community associated with adenoma stage of disease. Regardless of sample type and whether a full or select model was used, our Random Forest models consistently performed poorly. Yet, the step-wise decrease in diversity suggests that the adenoma-associated community is not normal but has changed subtly [Figure 1]. This change in diversity, at this early stage of disease, could be focal to the adenoma itself. One possible explanation is that how the host interacts with these subtle changes at early stages of the disease is what leads to a thoroughly dysfunctional community that is supportive of CRC genesis.

Within stool, common GI microbes were most consistently present in the top 10 genera or OTUs across studies [Figure 6]. Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be in the top 10 most important variables across the different studies for both adenoma and carcinoma [Figure 6]. These data suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* is not as important as how these bacteria interact with the changing resident community. Based on these observations, it is possible to hypothesize that small changes in community structure lead to new niches in which any one of the mouth-associated genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma stage of disease.

The colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with CRC Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S7E & F]. Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 most important variables of the full carcinoma stage models [Figure S7B-C & E-F]. The majority of the colon tissue-based results seem to be study specific with many of top 10 taxa being present only in a single study. Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models is worrying. The low bacterial biomass of tissue samples coupled with potential contamination could explain why the these results seem to be more sporadic than the stool results.

One important caveat to this study is that even though genera associated with certain species such as *Bacteroides fragilis* and *Streptococcus gallolyticus* subsp. *gallolyticus* were not identified, it does not necessarily mean that these specific species are not important in human CRC [@bfrag_Boleij_2014; @strep_Aymeric_2017]. Since we are limited in our aggregation of the data to the genus level, it is not possible to clearly delineate which species are contributing to overall disease progression. Our observations are not inconsistent with the previous literature on either *Bacteroides fragilis* or *Streptococcus gallolyticus* subsp. *gallolyticus*. As an example, the stool-based full community models consistently identified the genus *Bacteroides*, as well as OTUs that classified as *Bacteroides*, to be important model components across studies. This suggests that even though *Bacteroides* may not increase the RR of CRC and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC. Additionally, *Streptococcus gallolyticus* subsp. *gallolyticus* is a mouth-associated micobe, and the results from this study suggest that regardless sample type, mouth-associated genera are commonly associated with an increased RR for both adenoma and carcinoma stage of disease.

The associations between the microbiota and adenoma stage of disease are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes. None of the studies analyzed were properly powered to detect a 10% or lower change between cases and controls. The results within our meta-analysis suggest that a small effect size may well be the scope in which differences consistently occur between controls and adenoma stage of disease. Future studies investigating adenoma stage and the microbiota need to take power into consideration to reproducibly study whether the microbiota contributes to polyp formation. In contrast to adenoma stage of disease, our observations suggest that most studies analyzed have sufficient power to detect many changes in the carcinoma-associated microbiota because of large effect size differences between cases and controls [Figure 7].

@@

By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in CRC. Further, the data presented here suggests that mouth-associated microbes can gain a foothold within the colon and are are commonly associated with the greatest RR of having a carcinoma. No conclusive signal with these mouth-associated micobes could be detected for adenoma stage of disease. Our observations also highlight the importance of power and sample number considerations when investigating the microbiota and adenoma stage of disease due to the subtle changes in the community. Overall, associations of microbiota with the carcinoma stage of CRC are much stronger than those with the adenoma stage.

### 2018-01-26 submission/manuscript.tex
discussion and conclusion -- ada suggestions

\textbf{\emph{Select Genera Models can Recapitulate Whole Community
Models:}} Since specific genera increased RR for carcinoma over
diversity metrics we assessed whether the bacterial community was better
at classifying disease versus only a select group of genera based on RR
and P-value significance and tested this using two approaches. The first
approach used genus level data and tested whether there were any
differences in AUC between all genera and selected genera during
training on one study or testing on all other studies. The second
approach used OTU level data and tested whether there was a generalized
decrease in the 10-fold cross validation (CV) model AUC across studies
when comparing all OTUs versus only OTUs that taxonomically classified
to selected genera.

For the genera-based models the training set median AUC for model
classification was similar for select genera and full genera models, for
both tissue and stool studies {[}Figure S2-S3{]}. With respect to the
test sets, comprised of genera data from other studies, both all and
select genera models had a similar ability to detect adenomas or
carcinomas, with the select genera models performing better in some
instances {[}Figure S4-S6{]}. Conversely, the OTU-based models showed a
slight decrease in median AUC between the full and select models, with
one exception being carcinoma models for matched tissue {[}Figure 4 \&
5{]}.

In stool, the most common genera in the full genera-based models
belonged to genera such as \emph{Ruminococcus}, \emph{Bacteroides}, and
\emph{Roseburia} {[}Figure 6A \& B{]}. Although all mouth-associated
genera were present in carcinoma, none were present in the majority of
studies and \emph{Fusobacterium} was the only genus present in adenoma
{[}Figure 6A \& B{]}. For the full OTU-based models,
\emph{Ruminococcaceae} was present in the top 10 consistently for both
adenoma and carcinoma models while \emph{Roseburia} was only present in
many adenoma models and \emph{Bacteroides} was present in the
overwhelming majority of the carcinoma models {[}Figure 6C \& 6D{]}.

Unlike stool, the tissue full genera-based Random Forest models showed
no consistent representation of \emph{Ruminococcaceae},
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} in the top
10 across study {[}Figure S7{]}. The vast majority of the top 10 genera
and OTUs occured in a study specific manner for these tissue based
Random Forest models. For both the full genera and OTU Random Forest
models of adenoma and carcinoma, there appears to be very little overlap
in the top 10 most important variables between stool and tissue
{[}Figure 6 \& S7{]}. This discordance between stool and tissue also
applies to the mouth-associated genera with one noticeable skew being
that \emph{Fusobacterium} and \emph{Fusobacteriaceae} occur more often
in the top 10 of Random Forest models using matched versus unmatched
tissue samples {[}Figure S7B-C \& S7E-F{]}.

\textbf{\emph{CRC Studies are Underpowered for Detecting Small Effect
Sizes:}} Based on our reported results we assessed whether the studies
analyzed were realistically powered to identify small, medium, and large
scale differences between cases and controls. When assessing the power
of each study at different effect sizes the majority of studies, for
both adenoma and carcinoma, acheived 80\% power to detect a 30\% or
greater difference between groups {[}Figure 7A \& B{]}. No study that
was analyzed had the standard 80\% power to detect an effect size
difference that was equal to or below 10\% {[}Figure 7A \& B{]}. In
order to achieve a power of 80\%, for small effect sizes, studies used
in our meta-analysis would need to recruit over 1000 individuals for
each arm {[}Figure 7C{]}.

@@

Our study identifies clear differences in diversity, both at the
community level and for individual genera, that are present in patients
with and without CRC {[}Figure 1-3{]}. Although there was a step-wise
decrease in diversity from control to adenoma to carcinoma, this did not
translate into large effect sizes for the RR of lesion. Even though
mouth-associated genera increased the RR of carcinoma, they did not
consistently increase the RR of adenoma. These mouth-associated genera
are clearly important to carcinoma classification but our observations
suggest that accounting for other community members can increase the
ability of models to make predictions.

The data presented herein supports the importance of select genera for
carcinoma but not necessarily adenoma formation. Our observations show
that when only using the select genera to create models for carcinoma
the AUC are similar to using all genera or OTUs {[}Figure 4 \& 5{]}.
This suggests that an interplay between a select number of potentially
protective and exacerbating microbes, within the GI community, are
crucial for carcinoma formation. Importantly, it suggests that there may
be key members of the GI community that might be studied further to
potentially reduce the risk of carcinoma. Conversely, using the present
data, it is clear that new approaches may need to be used to identify
members of the community that are associated with adenoma. Regardless of
sample type and whether a full or select model was used our Random
Forest models consistently performed poorly. Yet, the step-wise decrease
in diversity suggests that the adenoma community is not normal but has
changed subtly {[}Figure 1{]}. This change in diversity, at this early
stage of disease, could be focal to the adenoma itself. One potential
hypothesis from these observations is that at early stages of the
disease, how the host interacts with these subtle changes is what
ultimately leads to a thoroughly dysfunctional community that is
supportive of CRC genesis.

Within stool, common GI microbes were most consistently present in the
top 10 genera or OTUs across study {[}Figure 6{]}. Changes in
\emph{Bacteroides}, \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
\emph{Roseburia} were consistently found to be disciminative across the
different studies for both adenoma and carcinoma {[}Figure 6{]}. This
data would suggest that whether the non-resident bacterium is
\emph{Fusobacteria} or \emph{Peptostreptococcus} is not as important as
how these bacteria interact with the changing resident community. Based
on these observations, it is possible to hypothesize that small changes
in community structure lead to new niches in which any one of the
mouth-associated genera can gain a foothold, exacerbating the initial
changes in community and facilitating the transition from adenoma to
carcinoma.

The tissue studies did not provide a clearer understanding of how the
microbiota may be associated with lesion. Generally, the full OTU-based
models of unmatched and matched {[}Figure S7E \& F{]} tissue samples
were concordant with stool showing that resident microbes were the most
prevalent in the top 10 across study. Unlike in stool,
\emph{Fusobacterium} was the only mouth-associated bacteria consistently
present in the top 10 of the full CRC models {[}Figure S7B-C \& E-F{]}.
The majority of the tissue results seem to be study specific with many
top 10 taxa being present only in a single study. Additionally, the
presence of genera associated with contamination, within the top 10 most
important variables for the genera and OTU models is worrying. The low
bacterial biomass of tissue samples coupled with potential contamination
could ultimately explain why the these results seem to be more sporadic
than the stool results.

One important caveat to this study is that even though specific genera
associated with species such as \emph{Bacteroides fragilis} and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
identified, it does not necessarily mean these specific species are not
important in human CRC {[}20,22{]}. Since we are limited in our
aggregation of the data, across study, to the genus level, it is not
possible to clearly delineate what specific species are contributing to
overall disease progression. Our observations are not inconsistent with
the previous literature on either \emph{Bacteroides fragilis} or
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus}. As an
example, for stool, the full RF models consistently identified the genus
\emph{Bacteroides} as well as OTUs that classified as \emph{Bacteroides}
to be important model components across studies. Suggesting that even
though \emph{Bacteroides} may not increase the RR of lesion and may not
vary in relative abundance, like \emph{Fusobacterium}, it is still
important in CRC. Additionaly, \emph{Streptococcus gallolyticus} subsp.
\emph{gallolyticus} is a mouth-associated micobe and our study clearly
shows that regardless of whether the sample is tissue or stool,
mouth-associated genera are commonly associated with an increased RR of
both adenoma and carcinoma.

The associations between the microbiota and adenoma are inconclusive, in
part, because many studies may not be powered effectively to observe
small effect sizes. None of the studies analyzed were properly powered
to detect a 10\% or lower change between cases and controls. A small
effect size may well be the scope in which differences consistently
occur in adenoma based on the results observed within our meta-analysis.
Future studies investigating adenoma and the microbiota need to take
power into consideration if we are to reproducibly study whether the
microbiota contributes to polyp formation. In contrast to adenoma, our
observations suggest that most studies analyzed have sufficient power to
detect many changes in carcinoma because of large effect size
differences between cases and controls {[}Figure 7{]}.

>>>>

\textbf{\emph{Select Genera Models can Recapitulate Whole Community
Models:}} Since specific genera increased RR for carcinoma over
diversity metrics we assessed whether the bacterial community was better
at classifying disease versus only a select group of genera. We selected
these genera based on their RR and P-value significance and used two
approaches to test this question. The first approach used genus level
data and tested for differences in AUC between all genera and selected
genera. A single study was used for training the model prior to testing
on all other studies and this was repeated for every study in the
meta-analysis. The second approach used OTU level data and tested for a
generalized decrease in the 10-fold cross validation (CV) model AUC
which is a common approach used to guard against over-fitting. This was
applied across study and the AUC of the all OTUs model was compared
against the model that used only OTUs that taxonomically classified to
selected genera.

For the first approach using the genera-based models, the training set
median AUC for model classification was similar for both the select
genera and full genera models, for both tissue and stool studies
{[}Figure S2-S3{]}. When analyzing the tests sets that were comprised of
genera data from other studies, both models had a similar ability to
detect individuals with adenomas or carcinomas, with the select genera
models performing better in some instances {[}Figure S4-S6{]}.
Conversely, the second approach that used OTU-based models showed a
slight decrease in median AUC between the full and select models, with
one exception to this generalization being the carcinoma models for
matched colon tissue {[}Figure 4 \& 5{]}.

In stool, the most common genera in the top 10 most imporant variables,
in models for the first approach, that used full genera-based models
were \emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia}
{[}Figure 6A \& B{]}. Regardless of sample type, mouth-associated genera
were present in models for the carcinoma stage of CRC {[}Figure 6A \&
B{]}. Yet, none were present in the majority of studies and
\emph{Fusobacterium} was the only genus present in the adenoma stage of
CRC {[}Figure 6A \& B{]}. For the second approach that utilized full
OTU-based models, \emph{Ruminococcaceae} was present in the top 10
consistently for both adenoma and carcinoma models while
\emph{Roseburia} was only present in many adenoma models and
\emph{Bacteroides} was present in the overwhelming majority of the
carcinoma models {[}Figure 6C \& 6D{]}.

Unlike the stool-based Random Forest models, the tissue-based models,
for the full genera from the first approach, showed no consistent
representation of \emph{Ruminococcaceae}, \emph{Ruminococcus},
\emph{Bacteroides}, and \emph{Roseburia} in the top 10 most important
model variables across study {[}Figure S7{]}. The vast majority of the
top 10 model variables for the genera- and OTU-based models using colon
tissue tended to be study specific. Further, there was very little
overlap in the top 10 important variables between adenoma and carcinoma
stage models, regardless of whether colon tissue or stool was used
{[}Figure 6 \& S7{]}. This discordance between stool and colon tissue
samples also applies to the mouth-associated genera with one noticeable
skew being that \emph{Fusobacterium} and \emph{Fusobacteriaceae} occur
more often in the top 10 of matched versus unmatched colon tissue Random
Forest models {[}Figure S7B-C \& S7E-F{]}. This suggests that either the
colon tissue microbiota is study and person dependent or that kit and/or
other types of contamination associated with low biomass samples may be
skewing the results.

\textbf{\emph{CRC Studies are Underpowered for Detecting Small Effect
Sizes:}} Next, we assessed how much confidence should be placed in the
reported outcomes from each individual study by calculating the ability
to detect a difference (power) and sample size needed for small, medium,
and large effect size differences between cases and controls. When
assessing the power of each study at different effect sizes the majority
of studies acheived 80\% power to detect a 30\% or greater difference
between groups {[}Figure 7A \& B{]}. No study that we analyzed had the
standard 80\% power to detect an effect size difference equal to or
below 10\% {[}Figure 7A \& B{]}. In order to achieve a power of 80\%,
for small effect sizes, studies used in our meta-analysis would need to
recruit over 1000 individuals for both the case and control arms
{[}Figure 7C{]}

@@

Our study identifies clear differences in diversity, both at the
community level and for individual genera, present in patients with and
without CRC {[}Figure 1-3{]}. Although there was a step-wise decrease in
diversity as disease progressed from control to adenoma to carcinoma,
this did not translate into large effect sizes for the RR of CRC. Even
though mouth-associated genera increased the RR of having a carcinoma,
they did not consistently increase the RR of having an adenoma.
Additionally, our observations suggest that by combining
mouth-associated and CRC protective microbes we can classify either
adenoma or carcinoma stage of disease as well as models that use the
full community.

The data presented herein support the importance of select genera for
carcinoma, but not necessarily adenoma, formation. The results that we
have presented show that the both genera and OTU select and full models,
for the carcinoma stage of CRC, had similar AUCs {[}Figure 4 \& 5{]}.
This suggests that an interplay between a select number of potentially
protective and exacerbating microbes within the GI community is crucial
for carcinoma formation. Importantly, it suggests that there may be key
members of the GI community that might be studied further to potentially
reduce the risk of carcinoma. Conversely, using the present data, it is
clear that new approaches may be needed to identify members of the
community associated with adenoma stage of disease. Regardless of sample
type and whether a full or select model was used, our Random Forest
models consistently performed poorly. Yet, the step-wise decrease in
diversity suggests that the adenoma-associated community is not normal
but has changed subtly {[}Figure 1{]}. This change in diversity, at this
early stage of disease, could be focal to the adenoma itself. One
possible explanation is that how the host interacts with these subtle
changes at early stages of the disease is what leads to a thoroughly
dysfunctional community that is supportive of CRC genesis.

Within stool, common GI microbes were most consistently present in the
top 10 genera or OTUs across studies {[}Figure 6{]}. Changes in
\emph{Bacteroides}, \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
\emph{Roseburia} were consistently found to be in the top 10 most
important variables across the different studies for both adenoma and
carcinoma {[}Figure 6{]}. These data suggest that whether the
non-resident bacterium is \emph{Fusobacteria} or
\emph{Peptostreptococcus} is not as important as how these bacteria
interact with the changing resident community. Based on these
observations, it is possible to hypothesize that small changes in
community structure lead to new niches in which any one of the
mouth-associated genera can gain a foothold, exacerbating the initial
changes in community and facilitating the transition from adenoma to
carcinoma stage of disease.

The colon tissue-based studies did not provide a clearer understanding
of how the microbiota may be associated with CRC Generally, the full
OTU-based models of unmatched and matched colon tissue samples were
concordant with stool samples showing that GI resident microbes were the
most prevalent in the top 10 most important variables across study
{[}Figure S7E \& F{]}. Unlike in stool, \emph{Fusobacterium} was the
only mouth-associated bacteria consistently present in the top 10 most
important variables of the full carcinoma stage models {[}Figure S7B-C
\& E-F{]}. The majority of the colon tissue-based results seem to be
study specific with many of top 10 taxa being present only in a single
study. Additionally, the presence of genera associated with
contamination, within the top 10 most important variables for the genera
and OTU models is worrying. The low bacterial biomass of tissue samples
coupled with potential contamination could explain why the these results
seem to be more sporadic than the stool results.

One important caveat to this study is that even though genera associated
with certain species such as \emph{Bacteroides fragilis} and
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
identified, it does not necessarily mean that these specific species are
not important in human CRC {[}20,22{]}. Since we are limited in our
aggregation of the data to the genus level, it is not possible to
clearly delineate which species are contributing to overall disease
progression. Our observations are not inconsistent with the previous
literature on either \emph{Bacteroides fragilis} or \emph{Streptococcus
gallolyticus} subsp. \emph{gallolyticus}. As an example, the stool-based
full community models consistently identified the genus
\emph{Bacteroides}, as well as OTUs that classified as
\emph{Bacteroides}, to be important model components across studies.
This suggests that even though \emph{Bacteroides} may not increase the
RR of CRC and may not vary in relative abundance, like
\emph{Fusobacterium}, it is still important in CRC. Additionally,
\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} is a
mouth-associated micobe, and the results from this study suggest that
regardless sample type, mouth-associated genera are commonly associated
with an increased RR for both adenoma and carcinoma stage of disease.

The associations between the microbiota and adenoma stage of disease are
inconclusive, in part, because many studies may not be powered
effectively to observe small effect sizes. None of the studies analyzed
were properly powered to detect a 10\% or lower change between cases and
controls. The results within our meta-analysis suggest that a small
effect size may well be the scope in which differences consistently
occur between controls and adenoma stage of disease. Future studies
investigating adenoma stage and the microbiota need to take power into
consideration to reproducibly study whether the microbiota contributes
to polyp formation. In contrast to adenoma stage of disease, our
observations suggest that most studies analyzed have sufficient power to
detect many changes in the carcinoma-associated microbiota because of
large effect size differences between cases and controls {[}Figure 7{]}.
